\section{Conclusion and outlook}
The EuBIC-MS Developers Meeting 2023 follows the successful previous winter schools and developers meetings.

In a follow-up survey, 50\% of the participants expressed their overall satisfaction with the meeting, with two-thirds of the survey respondents giving it a perfect score (5) and one-third a good score (4).
Overall, opinions varied, with a significant portion favoring either more leisure time or more organized activities, while a notable percentage found the existing balance to be satisfactory.
    
The EuBIC-MS Developers Meeting spanning a few days, witnessed substantial progress across all hackathon sessions, showcasing participants' impressive productivity and teamwork. The commitment of hackathon groups to ongoing collaboration and project completion holds promise for future scientific publications and improved computational mass spectrometric solutions to support proteomics and metabolomics.

Encouraged by the community’s enthusiastic support, the next EuBIC-MS Developers Meeting is scheduled to take place in February 2025, in Brixen, Italy \url{https://eubic-ms.org/events/2025-developers-meeting/}.
 
\section{Conflict of interest}
The authors declare no conflict of interest.
 
\section{Acknowledgements}
Sponsoring for the EuBIC-MS 2023 Developers Meeting was provided by \href{https://www.matrixscience.com/}{Matrix Science}, \href{https://www.msaid.de/}{MSAID} and \href{https://biognosys.com/}{Biognosys AG}. The poster price was sponsored by the \href{https://fgcz.ch}{Functional Genomics Center Zurich} of ETH Zurich and University of Zurich. The organization was supported by Congressi Stefano Franscini on Monte Verità, Switzerland. We would like to thank EuPA for their continuing support. We would like to thank all EuBIC-MS members who volunteered to help on all fronts, as well as all keynote speakers, hackathon organizers, and participants who contributed to the success of the Developers Meeting.

